[![](./images/lindens.jpg)](https://en.wikipedia.org/wiki/Tilia) An Application of LIME to a Random Forest

### 2019-03-01 presentations/graphics group/An Application of LIME to a Random Forest.Rmd
added code to folder
edits to presentation

1. Explanation of LIME 
2. Hamby bullet data
3. Applying LIME to the random forest
4. Issues with LIME and attempts at a solution
6. Conclusions and future work
]

@@

- Works with any black-box predictor

@@

Start with training data, complex model, and one prediction from the test data
1. Create perturbations based on training data.
2. Obtain a prediction for each perturbation using the complex model.
3. Compute a similarity score between the test observation and each perturbation. 
4. Perform feature selection to choose $k$.
5. Fit a linear regression (weighted by the similarity scores) to the perturbations with the $k$ chosen features (standardized).
  $$\mbox{prediction} \sim \mbox{chosen feature 1} + \cdots + \mbox{chosen feature k}$$
6. Use the coefficients from the regression to "explain" the predictions.

@@

- Sets of bullets from both “known” and “unknown” gun barrels were sent to firearm examiners around the world
- 240 total sets created
    - 10 barrels in a set
    - 35 bullets in a set 
        - 20 known (2 from the same barrel)
        - 15 unknowns (at least 1 and no more than 3 from each barrel)
- Examiners asked to use the known bullets to identify which barrels the unknown bullets were fired from

@@

- Want to know which were the driving variables for individual predictions
- Especially great when the model is wrong
- Variable importance scores provide global interpretations
- We want local interpretations

@@

.pull-left[
- quantile bins (default with 4 bins)
- equally spaced bins
]
.pull-right[
- normal approximation 
- kernel density estimation
]
.center[
```{r echo = FALSE, fig.height = 4.5, fig.width = 9}
limeout <- lime::lime(x = hamby173and252_train %>% select(ccf:sum_peaks), model = rtrees)
ggplot(hamby173and252_train, aes(x = ccf)) + 
  geom_histogram(bins = 30) + 
  geom_vline(xintercept = limeout$bin_cuts$ccf[[1]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[2]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[3]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[4]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[5]], color = "green", size = 1) +
  geom_vline(xintercept = hamby224_test %>% 
               filter(case == 1) %>% 
               select(ccf) %>%
               pull(), color = "blue", size = 1) +
  labs(x = "CCF", y = "Frequency") +
  theme(text = element_text(size = 18))
```
]
.center[<font size="4">Histogram of CCF from Training Data with LIME Quantile Bins (blue line is ccf value of case 1)</font>]

@@

#### Draw many samples from the estimated feature distributions to obtain the perturbations

- If the a binning estimation method is used, numeric features are converted to indicator variables.
    - 1 if in same bin as case 1
    - 0 o.w.

@@

$$\mbox{rf score}=\beta_0 + \beta_1 I[\mbox{ccf in same bin}] + \beta_2 I[\mbox{rough_cor in same bin}] +\cdots + \beta_9 I[\mbox{sum_peaks in same bin}]$$

@@

$$\mbox{rf score}=\beta_0 + \beta_1 I[\mbox{ccf in same bin}] + \beta_2 I[\mbox{cms in same bin}] + \beta_3 I[\mbox{matches in same bin}]$$

@@

We first applied lime using all default settings
- explanations did not make sense
- $R^2$ values were very low
- predictions from the simple model were poor

.center[<img src="./images/explain1.png" width=450>]
.center[<font size="4">Example where the random forest prediction is wrong. The explanations are from from 4 quantile bins.</font>]

@@

.center[<img src="./images/explain2.png" width=450>]
.center[<font size="4">This is the same case with explanations are from from 3 equally spaced bins.</font>]

@@

```{r echo = FALSE, message = FALSE}
source("../../code/helper_functions.R")
```

@@

- Equally spaced bins are dependent on the number of bins
- Tree based bins usually have the lowest MSEs
- rfscore tree based bins usually have the highest $R^2$ for the binning options
- All options results in low $R^2$ values
    - We think that the linear regression is not similar enough to a random forest to produce good explanations
- Explanations are relatively consistent across reps for first and second features
- Non-binning methods produce very different results

@@

- Apply LIME to explain a logistic regression model
    - We know how to interpret a logistic regression
    - Can compare to LIME explanations
- Run some sort of simulation to assess LIME explanations
- Adjust and compare weighting methods and the features selection methods
- Allow the number of bins to vary across features
- Use a tree as the simple model

>>>>

1. Explanation of LIME 
2. Hamby bullet data
3. Applying LIME to the random forest
4. Issues and attempts at a solution
6. Conclusions and future work
]

@@

- Works with any predictive model

@@

**Start with**: (1) training data, (2) complex model, (3) one prediction from the testing data
1. Create perturbations from training data
2. Input each perturbation into complex model to get predictions
3. Compute similarity scores between test observation and each perturbation 
4. Perform feature selection to choose $k$ features
5. Fit a simple and interpretable model weighted by the similarity scores to the perturbations
  $$\mbox{Example: } \mbox{prediction} \sim \mbox{feature 2} + \mbox{feature 3} \ \ \ \mbox{(with features standardized)}$$
6. Interpret the simple model to "explain" the predictions

@@

- Sets of bullets from both “known” and “unknown” gun barrels were sent to firearm examiners around the world
- 240 total sets created
    - 10 barrels used to create a set
    - 35 bullets in a set 
        - 20 knowns (2 from the same barrel)
        - 15 unknowns (at least 1 and no more than 3 from each barrel)
- Examiners asked to use the known bullets to identify which barrels the unknown bullets were fired from

@@

- Want to determine the driving variables for individual predictions
- Especially helpful when the model is wrong
- Variable importance scores provide global interpretations
- We want local interpretations

@@

.pull-left[
- quantile bins (default with 4 bins)
- equally spaced bins
]
.pull-right[
- normal approximation 
- kernel density estimation
]
.center[
```{r echo = FALSE, fig.height = 4.5, fig.width = 9}
limeout <- lime::lime(x = hamby173and252_train %>% select(ccf:sum_peaks), model = rtrees)
ggplot(hamby173and252_train, aes(x = ccf)) + 
  geom_histogram(bins = 30) + 
  geom_vline(xintercept = limeout$bin_cuts$ccf[[1]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[2]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[3]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[4]], color = "green", size = 1) +
  geom_vline(xintercept = limeout$bin_cuts$ccf[[5]], color = "green", size = 1) +
  geom_vline(xintercept = hamby224_test %>% 
               filter(case == 1) %>% 
               select(ccf) %>%
               pull(), color = "blue", size = 1) +
  labs(x = "CCF", y = "Frequency") +
  theme(text = element_text(size = 18))
```
]
.center[<font size="4">Histogram of CCF from training data with LIME quantile bins (blue line is ccf value of case 1)</font>]

@@

#### Draw many samples from the estimated feature distributions

- If a binning estimation method is used, numeric features are converted to indicator variables.
    - 1 if in same bin as case 1
    - 0 o.w.

@@

$$\hat{\mbox{rf score}}=\hat{\beta_0} + \hat{\beta_1} I[\mbox{ccf in same bin}] + \hat{\beta_2} I[\mbox{rough_cor in same bin}] +\cdots + \hat{\beta_9} I[\mbox{sum_peaks in same bin}]$$

@@

$$\hat{\mbox{rf score}}=\hat{\beta_0} + \hat{\beta_1} I[\mbox{ccf in same bin}] + \hat{\beta_2} I[\mbox{cms in same bin}] + \hat{\beta_3} I[\mbox{matches in same bin}]$$

@@

We first applied lime using all default settings.
- Explanations did not make sense
- $R^2$ values were very low
-Predictions from the simple model were poor

.center[<img src="./images/explain1.png" width=450>]
.center[<font size="4">Example where the random forest prediction is wrong. The explanations are from 4 quantile bins.</font>]

@@

.center[<img src="./images/explain2.png" width=450>]
.center[<font size="4">This is the same case with explanations from 3 equally spaced bins.</font>]

@@

```{r echo = FALSE, message = FALSE}
source("./code/helper_functions.R")
```

@@

- Equally spaced bins are dependent on the number of bins
- Tree based bins usually have the lowest MSEs
- `rfscore` tree based bins usually have the highest $R^2$ for the binning options
- All options results in low $R^2$ values
    - We think that the linear regression is not similar enough to a random forest to produce good explanations
- Explanations are relatively consistent across reps for first and second features
- Non-binning methods produce very different results compared to the binning methods

@@

- Apply LIME to a logistic regression model
    - We know how to interpret a logistic regression
    - Can compare to LIME explanations
- Run some sort of simulation to assess LIME explanations
- Adjust and compare weighting methods and the feature selection methods
- Allow the number of bins to vary across features
- Use a tree as the simple model
